---
title: "GDA_CaseStudy_Analyze"
author: "Fred"
date: "4/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Case Study Roadmap - Analyze

Ask Three questions will guide the future marketing program:

    1. How do annual members and casual riders use Cyclistic bikes dierently?
    2. Why would casual riders buy Cyclistic annual memberships? 
    3. How can Cyclistic use digital media to influence casual riders to become members? 
    
You will produce a report with the following deliverables: 

    1. A clear statement of the business task - Completed
    2. A description of all data sources used - Completed
    3. Documentation of any cleaning or manipulation of data - Completed
    4. A summary of your analysis - In progress
    5. Supporting visualizations and key findings
    6. Your top three recommendations based on your analysis

Guiding questions

    - How should you organize your data to perform analysis on it?
    - Has your data been properly formatted?
    - What surprises did you discover in the data?
    - What trends or relationships did you find in the data?
    - How will these insights help answer your business questions?

Key tasks

    1. Aggregate your data so it’s useful and accessible.
    2. Organize and format your data.
    3. Perform calculations
    4. Identify trends and relationships.

Deliverable

    - Here is a summary of the analysis of the Bike Share data from Jan, 2021 to March, 2021
    - The data cleaning has been performed in RStudio.
    - This analysis is for case study 1 from the Google Data Analytics Certificate (Cyclistic).

    

Setup environment in R:

    - Install required packages and library for data cleaning, transformation, and visualization.

```{r Install Packages, include=FALSE}
install.packages('tidyverse', repos = "http://cran.us.r-project.org")
install.packages("hms", repos = "http://cran.us.r-project.org")
install.packages("lubridate", repos = "http://cran.us.r-project.org")

### Load Necessary libraries
library(tidyverse)
library(hms)
library(lubridate)

options(scipen=999)  # turn off scientific notation like 1e+06
library(ggplot2)

```


Start of data cleaning and transformation:

    - Fist we need to load the csv files to a dataframe
    - Check data for consistency in reparation for cleaning and transoformation.
    - Inspect the new dataframe that has been created
    - Transform Necessary Columns for calculation
    - Create a ride length column
    - Remove unnecessary data.
    - Perform initial analysis of data
    
```{r Load csv files to dataframe, include=FALSE}
# path to folder that holds multiple .csv files

setwd("D:/Portfolio/GDA_Case_Study/Bike Share/data/divvy-tripdata/")  
file_list <- list.files()
csv <- lapply(file_list, read.csv)

all_trips <- do.call(rbind, csv)

```


```{r Inspect combined data, include=FALSE}
# Inspect the new dataframe that has been created

colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(qs_raw)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
#skim_without_charts(all_trips)
```


```{r trnasform Columns , include=FALSE}

#Transform data

all_trips$start_date <- as.Date(all_trips$started_at)
all_trips$end_date <- as.Date(all_trips$ended_at)
all_trips$month <- format(as.Date(all_trips$start_date), "%m")
all_trips$day <- format(as.Date(all_trips$start_date), "%d")
all_trips$year <- format(as.Date(all_trips$start_date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$start_date), "%A")

```


```{r Mutate Columns 3, include=FALSE}

#Create a ride_length column for calculation of trends. 
#Create a ride_length_hms column to display trip duration.

all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)


is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)


all_trips$ride_length_hms <- as.POSIXct(all_trips$ride_length, origin = "1960-01-01 00:00:00", "GMT")
all_trips$ride_length_hms <- format(all_trips$ride_length_hms, format = "%H:%M:%S")
all_trips$ride_length_hms <- as.hms(all_trips$ride_length_hms)

#mutate(weekday = wday(started_at, label = TRUE))
```


```{r Remove bad data, include=FALSE}

## Remove data with less than 0 ride_length

all_trips_v2 <- filter(all_trips, ride_length >= 0)

#all_trips_v2 <- all_trips %>%
#  filter(all_trips$ride_length <0)


# Inspect the structure of the columns
str(all_trips_v2)

```

Here is the statistical summary of the data:

```{r Initial Analysis, echo=FALSE}

# Descriptive analysis on ride_length (all figures in seconds)
#mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
#median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
#max(all_trips_v2$ride_length) #longest ride
#min(all_trips_v2$ride_length) #shortest ride

# You can condense the four lines above to one line using summary() on the specific attribute
summary(all_trips_v2$ride_length)

```

##### Compare members and casual users using mean

```{r Compare members and casual users, echo=FALSE}
# Compare members and casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
#aggregate(all_trips_v2$ride_length)
```

##### Compare members and casual users using median
```{r Compare members and casual users 2, echo=FALSE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
```

##### Compare members and casual users using max
```{r Compare members and casual users 3, echo=FALSE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
```

##### Compare members and casual users using min
```{r Compare members and casual users 4, echo=FALSE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```


```{r Fix week order, include=FALSE}
# Notice that the days of the week are out of order. Let's fix that.
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

```

##### See the average ride time by each day for members vs casual users
```{r See the average ride time by each day for members vs casual users 2, echo=FALSE}
# Now, let's run the average ride time by each day for members vs casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

```

#### Here is the data analysis visualization using a bar chart.
```{r echo=FALSE}

#data(all_trips_v2, package = "ggplot2")
all_trips_v2 %>%
    mutate(weekday = wday(started_at, label = TRUE)) %>% 
    group_by(member_casual, weekday) %>% 
    summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
    arrange(member_casual, weekday)  %>%
    ggplot(aes(x=weekday, y=number_of_rides, fill = member_casual)) +
    geom_bar(stat="identity", position = "dodge") + 
    scale_fill_brewer(palette="Dark2") +
    theme_minimal() + 
    labs(title = "Member and Casual Number of rides", x = "Day of Week", y = "Number of Rides", fill = "Member - Casual")


```



```{r Lets create a visualization for average duration, echo=FALSE}
# Let's create a visualization for average duration

all_trips_v2 %>%
    mutate(weekday = wday(started_at, label = TRUE)) %>% 
    group_by(member_casual, weekday) %>% 
    summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
    arrange(member_casual, weekday)  %>%
    ggplot(aes(x=weekday, y=average_duration, fill = member_casual)) +
    geom_bar(stat="identity", position = "dodge") + 
    scale_fill_brewer(palette="Dark2") +
    theme_minimal() + 
    labs(title = "Member and Casual average ride length", x = "Day of Week", y = "Average duration", fill = "Member - Casual")

```

#####
```{r eval=FALSE, include=FALSE}

```


